Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_natopen.Rmd
more prallel-comit-issues fixed

### Representativity in time and space {#rep-in-space-and-time-ndvin}
<!-- #### check here #### -->
For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 2881 natopen polygons are distributed across regions and major ecosystem types in the following way:

@@

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r, message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

**Where are the natopen sites?**
```{r where-are-natopen-sites, fig.cap="A map of Norway showing the location of the naturally open ecosystem sites. T12 = coastal meadows, T18 = flood plains, T2 = shallow soils, T20 = innfrysningsmark, T21 = sand dunes."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.natopen) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T12",
      "T18",
      "T2",
      "T20",
      "T21"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**
```{r histogram-sentinall-ndvin, fig.cap="Histogram of naturally open ecosystem polygon sizes in Norway."}
#summary(SentinelNDVI.natopen$area_meters)
hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.

**How does NDVI vary over the years?**
```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition.}
SentinelNDVI.natopen %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r NDVI-major-ndvin, warning=FALSE, fig.cap="NDVI values in major natopen ecosystem types in good condition."}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")
```
```{r}
SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
```

```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
```

@@

**How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="DVI variation across regions in natopen ecosystems (major types) in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition.}
SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

The plots suggest a slight decrease in NDVI as condition deteriorates. 
Overlaps between condition classes are generally very large.

<!-- #### continue here #### -->

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r balanced-data-ndvin, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by natopen ecosystem types in five regions in Norway. ", fig.height=12}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
  facet_grid(region ~ hovedtype)
```

@@

**Northern Norway**
```{r north-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Northern Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

**Central Norway**
```{r central-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Central Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

**Western Norway**
```{r west-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for natopen ecosystems in W-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

**Eastern Norway**
```{r east-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in E-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.natopen.cond.Sent.E)$coefficients
```

@@

**Southern Norway**
```{r south-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in S-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.natopen.cond.Sent.S)$coefficients
```

@@

# concatenate the three Satellite objects
allSatNDVI.natopen <- rbind(
  SentinelNDVI.natopen[, !names(SentinelNDVI.natopen) %in% "subtype"],
  ModisNDVI.natopen,
  LandsatNDVI.natopen[, !names(LandsatNDVI.natopen) %in% "column_label"]
)
# add variable for Satellite identity
allSatNDVI.natopen$Sat <- c(
  rep("Sentinel", nrow(SentinelNDVI.natopen)),
  rep("Modis", nrow(ModisNDVI.natopen)),
  rep("Landsat", nrow(LandsatNDVI.natopen))
)
allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat, levels = c("Sentinel", "Modis", "Landsat"))
   # levels(allSatNDVI.natopen$Sat)
```
Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
```{r ndvi-time-series-comparison-ndvin, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
# plot
allSatNDVI.natopen %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```

>>>>

### Representativity in time and space {#rep-in-space-and-time-ndvin}
For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 2881 natopen polygons are distributed across regions and major ecosystem types in the following way:

@@

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r message=FALSE, warning=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

**Where are the natopen sites?**
```{r where-are-natopen-sites, fig.cap="A map of Norway showing the location of the naturally open ecosystem sites. T12 = coastal meadows, T18 = flood plains, T2 = shallow soils, T20 = innfrysningsmark, T21 = sand dunes."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE, alpha=0) +
  tm_borders() +
  tm_shape(SentinelNDVI.natopen) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T12",
      "T18",
      "T2",
      "T20",
      "T21"
    ),
    title = "Major ecosystem type",
    size = 1.5
  ) +
  tm_layout(
    main.title = "Naturally open ecosystems below \nthe tree line \n(NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**
```{r histogram-sentinall-ndvin, fig.cap="Histogram of naturally open ecosystem polygon sizes in Norway."}
summary(SentinelNDVI.natopen$area_meters)
hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
abline(v = 100, lty = 2)
```

Natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.
The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.

**How does NDVI vary over the years?**
```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition.}
SentinelNDVI.natopen %>%
  filter(tilstand!="") %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Year")
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r NDVI-major-ndvin, warning=FALSE, fig.cap="NDVI values in major natopen ecosystem types in good condition."}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  ggtitle("NDVI for different ecosystem types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")
```

```{r}
SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
```

```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
  ggtitle("NDVI for different ecosystem sub-types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")

```

@@

**How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="DVI variation across regions in natopen ecosystems (major types) in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("NDVI for different ecosystems in good condition across regions") +
  labs(y = "NDVI values (Sentinel-2)", x = "Region")
```

@@

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition.}
SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  filter(tilstand != "") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
  ggtitle("NDVI for different ecosystems in different condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```
The plots suggest a slight decrease in NDVI as condition deteriorates. 
Overlaps between condition classes are generally very large.

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r balanced-data-ndvin, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by natopen ecosystem types in five regions in Norway. ", fig.height=12}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  filter(tilstand != "") %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey",) +
  facet_grid(region ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```

@@

**Northern Norway**
```{r north-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Northern Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Northern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

**Central Norway**
```{r central-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Central Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Central Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype) +
  facet_wrap(~hovedtype)
```

@@

**Western Norway**
```{r west-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for natopen ecosystems in W-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Western Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
  facet_wrap(~hovedtype)
```

@@

**Eastern Norway**
```{r east-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in E-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Eastern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
  facet_wrap(~hovedtype)
```
```{r}
summary(model.natopen.cond.Sent.E)$coefficients
```

@@

**Southern Norway**
```{r south-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in S-Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Southern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
  facet_wrap(~hovedtype)
```
```{r}
summary(model.natopen.cond.Sent.S)$coefficients
```

@@

# concatenate the three Satellite objects
allSatNDVI.natopen <- rbind(
  SentinelNDVI.natopen[, !names(SentinelNDVI.natopen) %in% "subtype"],
  ModisNDVI.natopen,
  LandsatNDVI.natopen[, !names(LandsatNDVI.natopen) %in% "column_label"]
)
# add variable for Satellite identity
allSatNDVI.natopen$Sat <- c(
  rep("Sentinel", nrow(SentinelNDVI.natopen)),
  rep("Modis", nrow(ModisNDVI.natopen)),
  rep("Landsat", nrow(LandsatNDVI.natopen))
)
allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat, levels = c("Sentinel", "Modis", "Landsat"))
   # levels(allSatNDVI.natopen$Sat)
```
Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
```{r ndvi-time-series-comparison-ndvin, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
# plot
allSatNDVI.natopen %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  ggtitle("NDVI across time") +
  labs(y = "NDVI values", x = "Year") +
  facet_wrap(~hovedtype, ncol = 1)
```
